Introduction

### 2020-08-14 text/abstract.Rmd
full edit (dennis), add PCAWG pathway analysis reference to intro, update word template to fix spacing before and after subsection headers

Cancer is a disease of cellular metabolism and other basic biochemical processes
common to all cells. However, molecular signatures remain largely
indistinguishable from tissue-of-origin, and general, unifying principles remain
obscure. To address this challenge, we reduced mutation data to disruptions in a
select set of pathways pointing to basic cell biology, from DNA replication to
cellular communication. Using dimension reduction techniques, we assigned tumor
samples into twelve classes distinct from their tissues-of-origin and largely
free of bias from mutational burden or clinical stage. We showed that these
tumor classes vary in prognosis by modeling relative risk of death in different
classes and cancer types. We identified class-specific mutations in different
tissues, demonstrating how tissue-specific signatures contribute to common
cellular phenotypes. Moreover, germline risk mutations involved in genome
stability, such as BRCA1 and Rb1, are equally distributed among our classes,
somewhat contrary to the expectation that such genes are hallmarks of molecular
subtypes. We also investigated pathways that distinguish between metastatic and
non-metastatic tumors, and found that most meaningful differences are
class-specific. Most metastatic enriched pathways from one class in metastatic
tumors are class-specific pathways from another class, suggesting phenotypic
convergence, and potentially highlighting key pathways to focus on
therapeutically for prevention of class-specific relapse. In summary, our
observations suggest that some genes act as drivers of genome instability in all
classes, while different sets of class-specific genes confer increased fitness
in a combinatorial fashion, eventually converging on one or more end-stage
disease phenotypes. Thus, we present a coherent picture of global tumor biology,
and explain how common cellular dysfunction might arise from tissue-specific
mutations.

>>>>

Cancer is a disease of cellular metabolism and other basic biochemical processes
common to all cells. However, molecular signatures remain largely
indistinguishable from tissue-of-origin, and general, unifying principles remain
obscure. To address this challenge, we reduced mutation data to disruptions in a
select set of pathways relevant to basic cell biology, from DNA replication to
cellular communication. Using dimension reduction techniques, we assigned tumor
samples into ten classes distinct from their tissues-of-origin and largely
free of bias from mutational burden or clinical stage. We showed that these
tumor classes vary in prognosis by modeling relative risk of death in different
classes and cancer types. We identified class-specific mutations in different
tissues, demonstrating that tissue-specific signatures contribute to common
cellular phenotypes. Moreover, driver gene mutations involved in replication fidelity and genome
stability, such as BRCA1 and Rb1, are equally distributed among our classes,
contrary to the expectation that such genes are avatars of molecular
subtype. We also investigated pathways that distinguish between metastatic and
non-metastatic tumors, and found that most meaningful differences are
class-specific. Many metastastasis enriched pathways from one class of
tumors are class-specific pathways from another class, suggesting phenotypic
convergence, and potentially highlighting key pathways to focus on
for prevention of class-specific relapse. In summary, our
observations suggest that some genes act as drivers of genome instability in all
classes, while different sets of class-specific genes confer increased fitness
in a combinatorial fashion, eventually converging on one or more end-stage
disease phenotypes. Thus, we present a coherent view of global tumor biology,
and explain how common cellular dysfunction might arise from tissue-specific
mutations.

### 2020-08-14 text/introduction.Rmd
full edit (dennis), add PCAWG pathway analysis reference to intro, update word template to fix spacing before and after subsection headers

Recent advances in technology have made the study of cancer possible at
unprecedented levels of detail and scale. Array and next-generation sequencing
technologies have made it possible to assay germline and somatic mutations,
expression profiles, DNA methylation, and copy number variations. To take
advantage of these tools several large consortia, including the cancer genome
atlas (TCGA) and pan-cancer analysis of whole genomes (PCAWG), sequenced large
numbers of tumors and collected data from multiple assays that could be analyzed
together, with the goal of integration and increased understanding of the
mechanisms of cancer. These efforts have focused on recognizing and quantifying
the importance of genes in cancer etiology as the result of selection processes,
or classification of cancers relative to common patterns of aberration in
multiple assays.

Considerable progress has been made analyzing these data. Statistical analyses
identify hundreds of global and tissue-specific cancer driver genes
[@deesMuSiC2012; @tamboreroOncodriveCLUST2013;
@lawrenceDiscoverySaturationAnalysis2014;
@kumarStatisticallyIdentifyingTumor2015; @tokheimEvaluatingEvaluationCancer2016;
@jiangWITER2019; @zhaoModelingPositive2019] using approaches aimed at detecting
when genes are mutated at a greater rate than expected due to chance. Based in
part on studies like these, it has been estimated that on the order of three
mutations in key oncogenes and/or tumor suppressors would be sufficient to
transform a normal cell to a cancerous state [@vogelsteinPathCancerThree2015].

Other studies aim at providing a comprehensive view of mutations, gene
expression and genomic signatures, with the express goal of understanding the
common themes of all cancer independent of tissue of origin. Such a view of
cancer as a malady affecting common cellular processes has long been a goal of
the field and is perhaps best characterized in essays by @weinberg2000; @hallmarks_II (see also
Discussion). The first of these studies, (the
“Pan-Cancer-12” analysis) identified 11 subtypes from 12 cancer types,
using integrative analysis with co-equal weighting of gene expression,
methylation, copy number and proteomics data
[@hoadleyMultiplatformAnalysis2014]. The principal finding was that
tissue-of-origin is the predominant driving factor, though a significant
proportion of tumors ($10\%$) could be reclassified independent of
tissue-of-origin. In a second study involving 33 cancer types and a much greater
number of tumors using a newer algorithm (iCluster)
[@shenIntegrativeClusteringMultiple2009], the authors identify 28 clusters that
could be further subdivided into organ specific groups, including
pan-gastrointestinal, pan-gynecological, pan-squamous, pan-gynecological/
squamous and pan-kidney [@hoadleyCellofOriginPatternsDominate2018].

More recently, attempts have been made to consider pan-cancer mutations and
other data outside the context of enrichment at the gene level. Horn _et al._
[@hornNetSigNetworkbasedDiscovery2018] introduce approaches to address the
sparseness of mutations, allowing the potential importance of a gene to be
influenced by the aggregate of mutations in its nearest network neighbors. This
method is powerful especially when applied to a small number (on the order of
hundreds) of tumors. One of the challenges in the field remains that beyond a
couple hundred common driver genes and mutations, there are likely thousands of
moderate effect genes that occur at such low frequency given tissue diversity
and low sample numbers that they are impossible to detect using positive
selection theory. Some groups have attempted to address this challenge by
applying machine learning to cancer data sets to discover groups of functionally
related genes as they interact with larger pathways and networks
[@kimPancancerAnalysisSomatic2018, @mourikis-ml-pathways, @moonlight]. These
studies have proven very effective at highlighting some potentially fundamental
disease phenotypes at the pathway level that cross cancers with different
origins at the cellular and tissue level.

In this study, we attempt to address these questions by focusing on discrete
molecular pathways as the unit of disruption in cancer. Our approach strips all
cell-type-specific information from the mutation data and generalizes gene-level
information to cell-biological pathway phenotypes. We use this heuristic to
evaluate all cancers and show that, surprisingly, tumors fall into categories of
common pathway disruptions with distinctive origins arising from tissue-specific
gene alterations and having unique prognosis in each cancer type.

>>>>

Recent advances in technology have made the study of cancer possible at
unprecedented levels of detail and scale. Array and next-generation sequencing
technologies have made it possible to assay germline and somatic mutations,
expression profiles, DNA methylation, and copy number variations. To take
advantage of these tools several large consortia, including the cancer genome
atlas (TCGA) and pan-cancer analysis of whole genomes (PCAWG), sequenced large
numbers of tumors and collected data from multiple assays that could be analyzed
together, with the goal of integration and increased understanding of the
mechanisms of cancer.

Considerable progress has been made analyzing these data. Statistical analyses
identify hundreds of global and tissue-specific cancer driver genes
[@deesMuSiC2012; @tamboreroOncodriveCLUST2013;
@lawrenceDiscoverySaturationAnalysis2014;
@kumarStatisticallyIdentifyingTumor2015; @tokheimEvaluatingEvaluationCancer2016;
@jiangWITER2019; @zhaoModelingPositive2019] using approaches aimed at detecting
when genes are mutated at a greater rate than expected due to chance. It has been estimated that on the order of three
mutations in key oncogenes and/or tumor suppressors would be sufficient to
transform a normal cell to a cancerous state [@vogelsteinPathCancerThree2015].

Other studies aim at providing a comprehensive view of mutations, gene
expression and genomic signatures, with the express goal of understanding the
common themes of all cancer independent of tissue of origin. Understanding cancer
as a disease of the cell has long been a goal of
the field and is perhaps best characterized in essays by @weinberg2000; @hallmarks_II (see also
Discussion). The first of these studies in the enomic era identified 11 subtypes from 12 cancer types,
using integrative analysis with co-equal weighting of gene expression,
methylation, copy number and proteomics data
[@hoadleyMultiplatformAnalysis2014]. The principal finding was that
tissue-of-origin is the predominant driving factor, though a significant
proportion of tumors (~10%$) could be reclassified independent of
tissue-of-origin. In a second study involving 33 cancer types and a much greater
number of tumors [@shenIntegrativeClusteringMultiple2009], the authors identify 28 clusters that
could be further subdivided into organ specific groups, including
pan-gastrointestinal, pan-gynecological, pan-squamous, pan-gynecological/
squamous and pan-kidney [@hoadleyCellofOriginPatternsDominate2018].

More recently, attempts have been made to consider pan-cancer mutations and
other data outside the context of enrichment at the gene level. @PCAWG-PATH carried out near-exhaustive analysis of pathway enrichment using whole genome data. Horn _et al._
[@hornNetSigNetworkbasedDiscovery2018] introduce approaches to address the
sparseness of mutations, allowing the potential importance of a gene to be
influenced by the aggregate of mutations in its nearest network neighbors. This
method is powerful especially when applied to a small number (on the order of
hundreds) of tumors. One of the challenges in the field remains that beyond a
couple hundred common driver genes and mutations, there are likely thousands of
moderate effect genes that occur at such low frequency--given tissue diversity--and low sample numbers that they are impossible to detect using positive
selection theory. Some groups have attempted to address this challenge by
applying machine learning to cancer data sets to discover groups of functionally
related genes as they interact with larger pathways and networks
[@kimPancancerAnalysisSomatic2018, @mourikis-ml-pathways, @moonlight]. These
studies have proven very effective at highlighting some potentially fundamental
disease phenotypes at the pathway level that cross cancers with different
origins at the cellular and tissue level.

In this study, we attempt to address these questions by focusing on discrete
molecular pathways as the unit of disruption in cancer. Our approach strips all
cell-type-specific information from the mutation data and equates gene-level
mutations to cell-biological pathway disruptions. We use this heuristic to
evaluate all cancers and show that, surprisingly, tumors fall into categories of
common pathway disruptions with distinctive origins arising from tissue-specific
gene alterations and having unique prognosis in each cancer type.
